---
title: Differential expression analysis
author: John Blischak
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options}
library("knitr")
opts_chunk$set(fig.align = "center", fig.width = 10)
```

## Setup

Load packages.

```{r packages}
library("edgeR")
library("ggplot2")
library("cowplot")
theme_set(theme_bw(base_size = 12))
```

Import the filtered counts matrix.

```{r input-counts}
counts <- read.table("../../data/mouse/counts-matrix-filtered.txt",
                     header = TRUE)
dim(counts)
head(counts)
```

Import the annotation data.

```{r import-anno}
anno <- read.table("../../data/mouse/annotation.txt", header = TRUE)
stopifnot(nrow(anno) == ncol(counts))
anno
```

## Differential expression analysis with edgeR

Convert to a DGEList.

```{r DGEList}
expr <- DGEList(counts = counts, group = anno$treatment,
                genes = rownames(counts))
```

Normalize using TMM.

```{r tmm}
expr <- calcNormFactors(expr)
```

Estimate dispersion.

```{r dispersion}
expr <- estimateCommonDisp(expr)
expr <- estimateTagwiseDisp(expr)
```

Perform test.

```{r exact-test}
de_test <- exactTest(expr)
topTags(de_test)
de_results <- topTags(de_test, n = nrow(counts), sort.by = "none")
de_results <- as.data.frame(de_results)
rownames(de_results) <- de_results$genes
de_results[["genes"]] <- NULL
fdr_cutoff <- 0.05
de_results$de <- de_results$FDR < fdr_cutoff
```

Using a FDR cutoff of `r fdr_cutoff * 100`%, `r sum(de_results$de)` of the `r nrow(counts)` genes are differentially expressed.

Save the results.

```{r save-de-results}
write.table(de_results, file = "../../data/mouse/de-results.txt",
            quote = FALSE, sep = "\t", col.names = NA)
```

## Visualize results

```{r visualize-de}
ma_plot <- ggplot(de_results, aes(x = logCPM, y = logFC, color = de)) +
  geom_point(alpha = 0.5) +
  scale_color_manual(values = c("black", "red")) +
  theme(legend.position = "none") +
  labs(title = "MA plot")
volcano_plot <- ma_plot %+% aes(x = logFC, y = -log10(PValue)) +
  xlim(-10, 10) + ylim(0, 100) +
  labs(title = "Volcano plot")
plot_grid(ma_plot, volcano_plot, labels = LETTERS[1:2])
```

## Session information

```{r session-info}
sessionInfo()
```
