---
title: Exploring the data
author: John Blischak
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options}
library("knitr")
opts_chunk$set(fig.align = "center", fig.width = 10)
```

## Setup

Load packages.

```{r packages}
library("edgeR")
library("ggplot2")
library("ggdendro")
library("cowplot")
theme_set(theme_bw(base_size = 12))
```

Import the counts matrix.

```{r input-counts}
counts <- read.table("../../data/mouse/counts-matrix.txt", header = TRUE)
dim(counts)
colnames(counts) <- sub("mESc.2i.RNA.", "", colnames(counts))
head(counts)
```

Create an annotation matrix.

```{r annotation}
anno <- strsplit(colnames(counts), split = "\\.")
anno <- data.frame(treatment = sapply(anno, function(x) x[1]),
                   replicate = sapply(anno, function(x) x[2]))
anno
```

## Filter genes

Remove genes with zero counts across all samples.

```{r remove-zeros}
zeros <- rowSums(counts) == 0
counts <- counts[!zeros, ]
```

There were `r sum(zeros)` genes with no observed counts across the `r ncol(counts)` samples.

Next remove lowly expressed genes.
Use TMM-normalized log~2~ counts per million (cpm).

```{r log-cpm}
normalized_lib_sizes <- calcNormFactors(counts)
log_cpm <- cpm(counts, log = TRUE,
               lib.size = colSums(counts) * normalized_lib_sizes)
```

There is a bimodal distribution of expression values.
A good cutoff to remove lowly expressed genes is a log~2~ cpm < 0 in all samples (red line).

```{r expression-distribution}
op <- par(mfcol = c(1, 3))
plot(density(log_cpm[, 1]), main = "Distribution of log2 cpm")
apply(log_cpm[, -1], 2, function(x) lines(density(x)))
abline(v = 0, col = "red")
boxplot(log_cpm, las = 3)
abline(h = 0, col = "red")
plot(rowMeans(log_cpm[, anno$treatment == "DMSO"]),
     rowMeans(log_cpm[, anno$treatment == "RA"]),
     main = "Mean log2 cpm", xlab = "DMSO", ylab = "RA")
abline(v = 0, h = 0, col = "red")
par(op)
```

Remove genes with mean log~2~ cpm < 0 in both the DMSO and RA-treated samples.

```{r remove-lowly-expressed}
lowly_expressed <- apply(log_cpm, 1, function(x) {
  mean(x[anno$treatment == "DMSO"]) < 0 & mean(x[anno$treatment == "RA"]) < 0
})
counts <- counts[!lowly_expressed, ]
```

Using this cutoff removes `r sum(lowly_expressed)` lowly expressed genes for a total of `r nrow(counts)` expressed genes, a reasonable number for a single cell type.

Now recalculate the log~2~ cpm with only the expressed genes.

```{r log-cpm-clean}
normalized_lib_sizes <- calcNormFactors(counts)
log_cpm <- cpm(counts, log = TRUE,
                lib.size = colSums(counts) * normalized_lib_sizes)
```

The lowly expressed genes have been removed, but still allowing for genes to be robustly expressed in only one of the two treatment conditions.

```{r expression-distribution-post-filtering}
op <- par(mfcol = c(1, 3))
plot(density(log_cpm[, 1]), main = "Distribution of log2 cpm")
apply(log_cpm[, -1], 2, function(x) lines(density(x)))
abline(v = 0, col = "red")
boxplot(log_cpm, las = 3)
abline(h = 0, col = "red")
plot(rowMeans(log_cpm[, anno$treatment == "DMSO"]),
     rowMeans(log_cpm[, anno$treatment == "RA"]),
     main = "Mean log2 cpm", xlab = "DMSO", ylab = "RA")
abline(v = 0, h = 0, col = "red")
par(op)
```

## PCA

Perform principal components analysis (PCA).
See [Zhang et al. 2009](http://www.ncbi.nlm.nih.gov/pubmed/19763933) for reference.

```{r calc-pca}
pca <- prcomp(t(log_cpm), retx = TRUE, center = TRUE, scale. = TRUE)
variances <- pca$sdev^2
explained <- variances / sum(variances)
barplot(explained * 100, main = "Variance explained by each PC", xlab = "PC",
     ylab = "Percent variance explained")
pca_df <- cbind(anno, pca$x)
```

As expected, the first two PCs explain most of the variation in the data.
The first PC separates the samples by treatment.
The second PC separates the RA-treated samples.
The hierarchical clustering (complete linkage of Euclidean distances) results agree with the PCA.

```{r plot-pca}
pca_plot <- ggplot(pca_df, aes(x = PC1, y = PC2, color = treatment,
                               shape = replicate)) +
  geom_point() +
  labs(x = sprintf("PC%d (%.2f%%)", 1, round(explained[1] * 100, 2)),
       y = sprintf("PC%d (%.2f%%)", 2, round(explained[2] * 100, 2)),
       title = "PCA")
hclust_result <- hclust(dist(t(log_cpm)))
dendro_plot <- ggdendrogram(hclust_result) +
  labs(title = "Hierarchical clustering")
plot_grid(pca_plot, dendro_plot, labels = LETTERS[1:2])
```

