---
title: Misregulation of H3K4me3
author: John Blischak
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options}
library("knitr")
opts_chunk$set(fig.align = "center", fig.width = 10)
```

There is a clear trend in the [metagene plot that H3K4me3 is misregulated across the gene body][body-meta].
The yeast with truncated Set1 have reduced H3K4me3 at the TSS and increased in the gene body.
How many genes are affected in this way?
To answer this question, I will calculate the difference between the mean rpm in bins 1-10 at the beginning of the gene body compared to bins 31-40, where the largest difference is seen the the metagene plot.
For each gene, I will test if this difference changes between yeast with the different versions of Set1.

[body-meta]: metagene.html#plot

## Setup

```{r load-packages, message=FALSE}
library("tidyr")
library("dplyr")
library("ggplot2")
library("cowplot")
theme_set(theme_bw(base_size = 12))
```

Read in the reads per million (rpm) data with the input subtracted.

```{r input-counts}
anno <- read.table("../../data/yeast/annotation.txt", header = TRUE, sep = "\t",
                   stringsAsFactors = FALSE)
counts_cpm_std <- read.table("../../data/yeast/counts-cpm-std.txt", header = TRUE, sep = "\t",
                   stringsAsFactors = FALSE) 
stopifnot(grep("H3K4me3", anno$sample, value = TRUE) == 
          grep("H3K4me3", colnames(counts_cpm_std), value = TRUE))
```

Calculate the difference.

```{r prepare-bins}
counts_long <- gather(counts_cpm_std, key = "sample", value = "cpm",
                      H3K4me3.aa762_1080.r1:H3K4me3.full_length.r2)
difference <- counts_long %>%
  group_by(sample, gene) %>%
  summarize(mean_tss = mean(cpm[bin %in% 1:10]),
            mean_body = mean(cpm[bin %in% 31:40]),
            difference = mean_tss - mean_body)
difference <- separate(difference, col = sample, into = c("ip", "set1", "replicate"),
                  sep = "\\.")
```

Remove genes where the difference is static across the samples (t-test throws an error).

```{r remove-static}
static_genes <- difference %>% group_by(gene) %>%
  summarize(variance = var(difference)) %>%
  filter(variance < 0.1) %>%
  select(gene) %>%
  unlist
difference <- difference %>% filter(!(gene %in% static_genes))
```

`r length(static_genes)` genes with zero variance in the difference were removed, leaving a total of `r length(unique(difference$gene))` for testing.

How much variation is there in these metrics?

```{r difference-variation}
difference_hist <- ggplot(difference, aes(x = difference)) +
  geom_histogram() +
  scale_x_log10() +
  labs(title = "Distribution of H3K4me3 difference",
       x = "difference (log10 transformed)")
compare_mean_plot <- ggplot(difference, aes(x = mean_tss, y = mean_body)) +
  geom_point() +
  scale_x_log10() + scale_y_log10() +
    labs(title = "Relationship between mean H3K4me3 in between TSS and gene body",
       x = "Mean rpm at gene body (log10 transformed)",
       y = "Mean rpm at TSS (log10 transformed)")
plot_grid(difference_hist, compare_mean_plot)
```

Peform the test.

```{r test-difference}
result <- difference %>% group_by(gene) %>%
  # filter(var(difference) > 1) %>%
  summarize(tstat = t.test(difference[set1 == "full_length"],
                           difference[set1 == "aa762_1080"])$statistic,
            pval = t.test(difference[set1 == "full_length"],
                           difference[set1 == "aa762_1080"])$p.value)
result$facet <- cut(result$tstat, breaks = quantile(result$tstat),
                    include.lowest = TRUE)
```

There is a clear enrichment of signal in the p-value distribution.

```{r pval-hist}
ggplot(result, aes(x = pval)) +
  geom_histogram() +
  labs(title = "p-value distribution",
       x = "p-value") +
  geom_vline(x = 0.05, col = "red")
n_sig <- sum(result$pval < .05)
n_sig_fdr <- sum(p.adjust(result$pval) < .05)
```

`r n_sig` of the total `r nrow(result)` genes tested had a significant difference in the H3K4me3 distribution.
However, after FDR correction for multiple testing, `r n_sig_fdr` genes were significant.
This is likely due to the small sample size.

## Plot metagene split by differences in H3K4me3 misregulation

Using the t-statistic from the above test, plot the distribution of H3K4me3 for genes with varying levels of evidence for misregulation.

```{r prepare-metagene-data}
counts_long <- merge(counts_long, result, by = "gene")

metagene <- counts_long %>%
  group_by(sample, bin, facet) %>%
  summarize(mean = mean(cpm),
            sem = sd(cpm) / sqrt(n()))
metagene$bin <- as.numeric(metagene$bin)
metagene$sample <- factor(metagene$sample,
                          levels = c("H3K4me3.full_length.r1",
                                     "H3K4me3.full_length.r2",
                                     "H3K4me3.aa762_1080.r1",
                                     "H3K4me3.aa762_1080.r2"),
                          labels = c("H3K4me3 full length Rep 1",
                                     "H3K4me3 full length Rep 2",
                                     "H3K4me3 aa762-1080 Rep 1",
                                     "H3K4me3 aa762-1080 Rep 2"))
```

Plot the metagene result.

```{r gene-body-metagene-tstat}
ggplot(metagene, aes(x = bin, y = mean)) +
  geom_line(aes(color = sample)) +
  geom_ribbon(aes(ymin = mean - sem, ymax = mean + sem, fill = sample),
              alpha = 0.25) +
  labs(x = "", y = "mean rpm (+/- sem)",
       title = "Distribution of H3K4me3 across gene body") +
  facet_wrap(~facet) +
  theme(legend.position = "bottom")
```

Each quantile contains ~`r table(result$facet)[4]` genes.

```{r num-per-facet}
table(result$facet)
```

## Session information

```{r session-info}
sessionInfo()
```
